\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices

### 2016-12-22 nobuild/jss_revision/response.tex

On the other hand, all.equal(hs, true.hess) is comparing a true analytic
calculation with a first order simple difference aproximation using the true
gradient function:
\begin{verbatim}
max(abs(  obj$hessian(P)
     - binary.hess(P, data=binary, priors=priors, order.row=order.row)))
[1] 2.786891e-06
\end{verbatim}

which might also be mentioned in the text. (Really just for exposition
purposes, after all, it is almost the main purpose of the package.)
  \end{revQuote}

\begin{response}
  Good point.  I revised that section to explain that evaluations of
  the function and gradient have to be identical to the true values,
  and I even replaced \func{all.equal} with \func{identical} to
  highlight that fact. I then explained that the
  \class{sparseHessianFD} calculation of the Hessian will not be
  identical to the true value, and report the mean relative
  difference.  I chose the mean relative difference over maximum
  absolute difference because the former is what the \func{all.equal}
  function uses to test if two matrices are equal within numeric tolerance.
\end{response}

\item \label{it:complex1}\begin{revQuote}
I think it would be instructive to add some of the following comparisons
with the above on p12.  The package numDeriv function hessian by default
does a second order Richardson approximation using the true function value
approximation. This involves a very large number of function evaluations in
an attempt to obtain some accuracy, but the accuracy is limited by being an
approximation of a second difference:
\begin{verbatim}
Hessian Richardson

@@

binary.grad(P + 0+1i, data=binary, priors=priors, order.row=order.row)

returning a complex result. (This does not rule out all possible problems.)

As I recall, sums, multiplication, and exponentiation are all complex
analytic, so it would not be too surprising if the example in the paper is
too, but I have not analyzed that. However, based on the result being very
good, it seems highly likely.
  \end{revQuote}

  \begin{response}
    See response to Item \ref{it:complex2}
  \end{response}

\item\label{it:complex2} \begin{revQuote}
A possible extension to the package would be to implement the complex
method in the sparse code. The function numDeriv:::jacobian.default
implements both simple and complex, so provides a good comparison of the
necessary (non-sparse) computation.

@@

  \end{revQuote}

\begin{response}

Here I am combining my responses to items \ref{it:complex1} and
\ref{it:complex2} together.

  The complex step function is now implemented in the package, and is
  introduced in Section 2.5 of the paper.  You
  are correct that ``most R primitives work with complex numbers,''
  but there are many commonly used functions that do not, including 
  \func{log1p}, \func{gamma}, and the probability distribution
  functions. This limitation on the kinds of functions that can be
  implemented easily in R, combined with the requirement that the function be
  holomorphic, and my own general unfamiliarity with the method, make
  me hesitant to promote it too much in the paper. But it is a nifty
  trick, and including it in the package was a good idea (thanks!).

In terms of accuracy, the paper includes comparisons between the true
Hessian and the two \class{sparseHessianFD} methods (finite differences
and complex step). I am not comparing accuracy with \pkg{numDeriv},
because that really is not the point of this paper.  However, for the
timing comparisons in Table 4, I did
change the baseline \pkg{numDeriv} method from \func{hessian} with the
\method{Richardson} method, to \func{jacobian} with the
\method{complex} method. Removing the second-order approximation saves
a lot of time, and lets me compute Table 4 without resorting to
parallel computation. That will simplify attempts at replication, and
removes another variable that might influence results.

@@

\begin{response}
  Fixed.  Thanks.
\end{response}

\item \begin{revQuote}
 The file replication.R does not set the RNG seed. This may not be too
important if only times are generated, but will be if resulting values are
included.

  \end{revQuote}

\begin{response}
  I added a \func{set.seed} statement near the top of \filename{replication.R}.
\end{response}

@@

(My laptop is a  Intel(R) Core(TM) i5-3337U CPU @ 1.80GHz,  4GB RAM, SSD
swap, running Mint variant of Ubuntu 14.04.2 LTS.)

\begin{verbatim}
Code and output removed
\end{verbatim}

  \end{revQuote}

  \begin{response}
    I agree.  See my response to Item \ref{it:complex2}.  I include the
    times from the 
    \method{complex} method instead of \method{simple} because of the
    improved accuracy.  I considered including both times in the
    table, but that would have cluttered the table, there was very
    little difference in the times between them, and the comparison
    between two \pkg{numDeriv} methods is not the point of this
    paper. But I completely agree that either \func{jacobian} method
    is a more valid comparison than using \func{hessian}.
  \end{response}

@@

 - Consider implementing a complex step method, and do a comparison.

 - Assess the difference when multiple CPUs are used. (run.par == FALSE vs
run.par == TRUE)

  \end{revQuote}

  \begin{response}
These are all good suggestions.  Let me address them point by point.
    \begin{itemize}
    \item It really is all about the sparsity pattern, and
      specifically, the how few partitions are needed to group the
      variables.  Of course, it will be hard to keep the number of
      partitions small if there are too many non-zero elements, so the
      sparsity itself has some effect.  The arrangement of the non-zeros is
      related to the ordering of the variables, but the permutation
      handles that during the initialization of the
      \class{sparseHessianFD} argument.  For example, variables
      arranged in a way that the Hessian has a block-arrow pattern are
      permuted to get a Hessian with a banded pattern instead.  This
      is all discussed in the ``Partitioning the
      variables'' section of the paper.
\item I have no idea how to do this in a generalizable way.  I think
  it has a lot to do with the architecture of the computer
  itself. Figure 4 does break down the components of the computation
  into the time it takes to reserve memory (initialization), partition
  the variables, and compute gradients themselves.

I do not want to ignore this suggestion, but I am simply not clear on
what is being asked.  I am open to more specific suggestions.

\item The complex step method is now supported, and comparisons of
  accuracy and computation time are now included in the paper.

\item Parallel computing on a shared memory architecture really
  muddles up timing comparisons.  The \emph{only} reason I generated
  Table 4 in parallel was because of the \pkg{numDeriv} \func{hessian}
  function. Now that \func{hessian} was replaced with \func{jacobian},
  the baseline measures from \pkg{numDeriv} run fast enough that I can
  generate Table 4 in serial.

Your point is well taken that timing comparisons can be affected by
whether computation is being run on multiple processors, but the
``serialization'' of Table 4 makes the issue moot.

    \end{itemize}

### 2016-12-22 nobuild/jss_revision/sparseHessianFD.Rnw
more work on revision and response

\subsection[An example]{An example}

@@

@@

were computed by the same functions

@@

In short, the complex step method is available to users who want
improved accuracy and speed, and know what they are doing.

>>>>

\subsection[An example]{An example}\label{sec:exHess}

@@

The log posterior density in Equation \ref{eq:LPD} is holomorphic, so we
can estimate its Hessian using the complex step method, and compute
the mean relative difference from the true Hessian.

<<>>=
obj2 <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
  rows=pattern[["rows"]], cols=pattern[["cols"]],
  complex=TRUE,
  data=binary, priors=priors, order.row=order.row)
hs2 <- obj2$hessian(P)
mean(abs(hs2-true.hess))/mean(abs(hs2))
@

@@

In short, the complex step method can be much more accurate than
finite differencing, but it comes with theoretical and implementation
restrictions that may limit its universality.

@@

### 2016-12-22 nobuild/jss_revision/sparseHessianFD.tex
more work on revision and response

\begin{Schunk}
\begin{Sinput}
R> hs <- obj$hessian(P)
R> mean(abs(hs-true.hess))/mean(abs(hs))
\end{Sinput}
\begin{Soutput}
[1] 1.1217e-08
\end{Soutput}
\end{Schunk}

@@

were computed by the same functions

>>>>

\begin{Schunk}
\begin{Sinput}
R> hs <- obj$hessian(P)
R> mean(abs(hs-true.hess))/mean(abs(hs))
\end{Sinput}
\begin{Soutput}
[1] 1.12e-08
\end{Soutput}
\end{Schunk}

@@

The log posterior density in Equation \ref{eq:LPD} is holomorphic, so we
can estimate its Hessian using the complex step method, and compute
the mean relative difference from the true Hessian.

\begin{Schunk}
\begin{Sinput}
R> obj2 <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
+   rows=pattern[["rows"]], cols=pattern[["cols"]],
+   complex=TRUE,
+   data=binary, priors=priors, order.row=order.row)
R> hs2 <- obj2$hessian(P)
R> mean(abs(hs2-true.hess))/mean(abs(hs2))
\end{Sinput}
\begin{Soutput}
[1] 6.75e-18
\end{Soutput}
\end{Schunk}

In short, the complex step method can be much more accurate than
finite differencing, but it comes with theoretical and implementation
restrictions that may limit its universality.
